---
title: "XGboost for the feature importance for each potentially causally linked metabolic measure"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  rmdformats::html_clean:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
---

# Init

```{css, echo = FALSE}
.main-container {
  max-width:1920px !important;
}
```

```{r options, include = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, results='asis', cache=FALSE, warning=FALSE, collapse = TRUE, error = TRUE)
```


# Library

```{r}
library(magrittr)
```

# Devtools

```{r devtools}
devtools::load_all(quiet = TRUE)
```

## Command line arguments

```{r}
args <- list(optparse::make_option("--results", default = "nested_xgboost_cox_cv_.*-20240325_113000.rds")) %>% 
  optparse::OptionParser(option_list = .) %>% 
  optparse::parse_args()
```

```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% knitr::kable()
```

## Function factories

```{r}
mysave <- mysavefactory()
```

## Plot settings

```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(panel.border = ggplot2::element_rect(colour = "black", fill = NA),
                   panel.background = ggplot2::element_rect(fill = NA, color = NA),
                   plot.background = ggplot2::element_rect(fill = NA, color = NA), 
                   axis.text = ggplot2::element_text(colour = "black", size = 10)) } %>%
  ggplot2::theme_set()
```

# Load previously run nested XGboost models

```{r}
result_files <- list.files(path = "rds", pattern = args$results, full.names = TRUE)
```

```{r}
result_files %>%
  tibble::as_tibble() %>% 
  DT::datatable()
```

```{r Read result files}
results <- result_files %>% 
  purrr::map(nested_xgboost_cox_cv_load, .progress = TRUE) %>%
  rlang::set_names(., glue::glue("model_{seq_along(.)}"))
```

# First plot panel

## Harrell's statistic

```{r calculate harrells statistic}
df_harrell <- results %>%
  purrr::map_df(~purrr::map_df(.x, nested_xgboost_cox_cv_harrell, .id = "inner"), .id = "outer") %>%
  dplyr::summarize(index = mean(index), .by = outer)
```

```{r}
df_harrell %>% dplyr::summarize(mean = mean(index), sd = sd(index)) %>% knitr::kable()
```


```{r}
df_harrell %>% DT::datatable()
```

```{r}
{ panel_harrell <- { df_harrell %>%
                 ggplot2::ggplot(ggplot2::aes(x = index, fill = outer, color = outer)) +
                 ggplot2::geom_histogram(binwidth = 0.005, boundary = 0.005) +
                 ggplot2::scale_fill_grey(start = 0, end = 0.9) +
                 ggplot2::scale_color_grey(start = 0, end = 0.9) +
                 ggplot2::scale_y_continuous(name = NULL, expand = c(0, 0)) +
                 ggplot2::scale_x_continuous(name = "Predictive power of nested survival models using Harrell's C score",
                                             expand = c(0, 0),
                                             limits = ~ c(round(20*min(abs(.x)), 0)/20, 1),
                                             labels = scales::percent) +
                 ggplot2::ggtitle('C') +
                 ggplot2::theme(legend.position = "none",
                                panel.grid.major = ggplot2::element_blank(),
                                panel.grid.minor = ggplot2::element_blank(),
                                plot.margin = ggplot2::margin(1, 5, 1, 1, "mm")) } %>% 
    grob_resize_panels(., heights =  20, width = 150) } %>%
  mysave(name = "xgboost_harrell", width = 10, height = 3)
```

## Feature gain

```{r calculate model gain}
df_gain <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}") %>%
  purrr::map_df(~xgboost::xgb.importance(colnames(.x$matrix), .x$model), .id = "model")
```

```{r}
df_gain %>%
  dplyr::summarize(Gain = sum(Gain), .by = Feature) %>%
  my_add_names(by = "Feature") %>%
  dplyr::mutate(Desc = ifelse(Feature == "PREVAL_MACE_PAD", "Peripheral artery disease", Desc)) %>% 
  dplyr::arrange(desc(Gain)) %>%
  DT::datatable(caption = "Feature gain")
```

### Features better than randomness

```{r number of better than random features}
n_features <- df_gain %>%
  dplyr::summarize(Gain = sum(Gain), .by = Feature) %>% 
  dplyr::arrange(desc(Gain)) %>%
  dplyr::mutate(position = dplyr::row_number() - 1) %>% 
  dplyr::filter(stringr::str_detect(Feature, "rnorm_")) %>%
  dplyr::pull(position) %>%
  min()
```

Observed number of `r n_features` better than random features.

### XGBoost importance

```{r}
df_gain_scaled <- df_gain %>%
  dplyr::mutate(model = stringr::str_replace(model, "_[0-9]*$", "")) %>% 
  dplyr::mutate(Gain_scaled = Gain/length(results))
```

```{r}
df_importance_sum <- df_gain_scaled %>%
  dplyr::summarize(sum = sum(Gain_scaled), .by = Feature) %>%
  dplyr::arrange(desc(sum)) %>%
  dplyr::filter(dplyr::row_number() <= n_features) %>% 
  my_add_names(by = "Feature") %>%
  dplyr::mutate(Desc = ifelse(is.na(Desc) | nchar(Desc) == 0, Feature, Desc)) %>%
  dplyr::mutate(Feature = forcats::fct_rev(factor(dplyr::row_number(), labels = Feature))) %>% 
  dplyr::mutate(Desc = forcats::fct_rev(factor(dplyr::row_number(), labels = Desc)))
```

```{r}
df_importance <- df_gain_scaled %>%
  dplyr::inner_join(df_importance_sum, by = dplyr::join_by(Feature)) %>%
  dplyr:::arrange(sum) %>% 
  dplyr::mutate(Desc = ifelse(stringr::str_detect(Feature, "GUT_|Phylum_|Class_|Order_|Family_|Genus_|Species_"), glue::glue("*{Desc}*"), as.character(Desc))) %>%
  dplyr::mutate(Desc = ifelse(Desc == "Free cholesterol to total lipids ratio in very large VLDL",
                              "Free cholesterol ratio in very large VLDL",
                              Desc)) %>% 
  dplyr::mutate(Desc = factor(dplyr::row_number(), labels = Desc)) %>%
  dplyr::mutate(panel = dplyr::case_when(stringr::str_detect(Feature, "Phylum_|Class_|Order_|Family_|Genus_|Species_") ~ "Taxa",
                                         stringr::str_detect(Feature, "NMR_") ~ "Metab",
                                         TRUE ~ "Other") %>% factor(levels = c("Other", "Taxa", "Metab")))  
```

```{r}
c_importance_order <- df_importance_sum %>% dplyr::pull(Feature) %>% as.character()
```

```{r, include = FALSE}
df_gain_scaled %>%
  dplyr::summarize(sum = sum(Gain_scaled), .by = Feature) %>%
  dplyr::arrange(desc(sum)) %>%
  readr::write_tsv("xgboost_importance.tsv")
```

### Importance plot

```{r}
df_subset <- df_importance %>%
  dplyr::filter(sum > 0.03 | stringr::str_detect(Feature, "NMR_|GUT_|Phylum_|Class_|Order_|Family_|Genus_|Species_|shannon")) %>%
  dplyr::select(Feature) %>%
  unique
```

```{r}
panel_heights <- get_distinct_values_by_group(df = df_importance %>% dplyr::inner_join(df_subset, by = dplyr::join_by(Feature)),
                                              col = Desc,
                                              group_by = panel,
                                              multiplier = 4,
                                              list = TRUE)
```

```{r}
{ panel_gain <- { df_importance %>%
                    dplyr::inner_join(df_subset, by = dplyr::join_by(Feature)) %>%
                    ggplot2::ggplot(ggplot2::aes(x = Desc, y = Gain_scaled, fill = model, color = model)) +
                 ggplot2::facet_wrap(~panel, ncol = 1, scales = "free") +
                 ggplot2::geom_bar(stat = "identity") +
                 ggplot2::coord_flip() +
                 ggplot2::scale_color_grey(start = 0, end = 0.9) +  
                 ggplot2::scale_fill_grey(start = 0, end = 0.9) +
                 ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) +
                 ggplot2::ggtitle('A') +
                 ggplot2::scale_y_continuous(name = "Covariate's\ngain",
                                             expand = c(0, 0),
                                             limits = ~ ifelse(max(.x) > 0.03, 0.4, 0.03) %>% c(0, .),
                                             breaks = ~ ifelse(max(.x) > 0.03, 0.2, 0.01) %>% seq(0, 1, .),
                                             labels = scales::percent) +
                 ggplot2::theme(strip.background = ggplot2::element_blank() ,
                                strip.text.x = ggplot2::element_blank(),
                                panel.spacing = ggplot2::unit(2, "mm"),
                                plot.margin = ggplot2::margin(0, 0, 0, 0, "mm"),
                                legend.position = "none",
                                axis.text.y = ggtext::element_markdown(),
                                panel.grid.major = ggplot2::element_blank(),
                                panel.grid.minor = ggplot2::element_blank()) } %>%
    grob_resize_panels(., heights = panel_heights, width = 30) } %>%
  mysave(name = "xgboost_importance", width = 6, height = 10)
```

## Shapley values

```{r}
calculate_shaps <- function(x, vars) {
  object <- shapviz::shapviz(x$model, X = x$data, X_pred = x$matrix)
  S <- shapviz::get_shap_values(object) %>%
    tibble::as_tibble() %>%
    tibble::add_column(ids = x$ids)
  X <- shapviz::get_feature_values(object) %>%
    tibble::as_tibble() %>%
    tibble::add_column(ids = x$ids)
  list(X = X, S = S)
}
```

```{r calculate shapley values}
shap_interactions <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}") %>%
  purrr::map(calculate_shaps)
```

## SHAP beeswarm


```{r}
S_subset <- shap_interactions %>%
  purrr::map_df(~.x$S[, c_importance_order %union% "ids"], .id = "model") %>% 
  dplyr::summarize(dplyr::across(dplyr::one_of(c_importance_order), median), .by = ids) %>%
  tibble::column_to_rownames("ids") %>% 
  as.matrix()
```

```{r}
X_subset <- shap_interactions %>%
  purrr::map_df(~.x$X[, c_importance_order %union% "ids"], .id = "model") %>% 
  dplyr::summarize(dplyr::across(dplyr::one_of(c_importance_order), median), .by = ids) %>% 
  dplyr::mutate(dplyr::across(dplyr::one_of(c_importance_order), shapviz:::.min_max_scale)) %>%
  tibble::column_to_rownames("ids") %>% 
  as.matrix()
```

```{r, include = FALSE}
df_shap <- readRDS("rds/df_shap-20241105_124917.rds")
```

```{r}
df_shap <- as.data.frame.table(S_subset, responseName = "value") %>%
  dplyr::mutate(Feature = forcats::fct_rev(factor(Var2, levels = c_importance_order))) %>% 
  dplyr::mutate(color = as.data.frame.table(X_subset)$Freq) %>%
  tidyr::drop_na()
mysave(df_shap)
```


```{r}
my_x_breaks <- function(x, limits = FALSE) {
  diff <- max(x) - min(x)
  step <- dplyr::case_when(diff < 1 ~ 0.1,
                           TRUE ~ 1.0)
  seq(-10, 10, step)
}

```


```{r}
panel_b <- { df_shap %>%
               dplyr::inner_join(df_subset, by = dplyr::join_by(Feature)) %>%
               dplyr::left_join(df_importance %>% dplyr::select(Feature, Desc, panel, sum) %>% dplyr::distinct(),
                                by = dplyr::join_by(Feature)) %>% 
               ggplot2::ggplot(ggplot2::aes(x = value, y = Desc)) +
               ggplot2::facet_wrap(~panel, ncol = 1, scales = "free") +
               ggplot2::geom_vline(xintercept = 0, color = "darkgray") +
               ggplot2::geom_point(ggplot2::aes(color = color),
                                   position = shapviz:::position_bee(width = 0.4, adjust = 1.0),
                                   size = 0.01,
                                   alpha = 0.2) +
               ggplot2::scale_colour_gradientn(colours = c("blue", "red"),
                                               breaks = c(0, 1),
                                               labels = c("low", "high")) +
               ggplot2::ggtitle('B') +
               ggplot2::scale_x_continuous(limits = ~ ifelse(max(abs(.x)) > 0.13, 1.3, 0.13) *  c(-1, 1),
                                           breaks = my_x_breaks) +
               ggplot2::labs(x = "Covariate's impact\nto CAD", y = ggplot2::element_blank(), color = "Covariate\nvalue") +
               ggplot2::theme(strip.background = ggplot2::element_blank() ,
                              panel.background = ggplot2::element_blank(),
                              plot.margin = ggplot2::margin(0, 0, 0, 0, "mm"),
                              strip.text.x = ggplot2::element_blank(),
                              axis.text.y = ggplot2::element_blank(),
                              panel.spacing = ggplot2::unit(2, "mm")) } %>%
  grob_resize_panels(., heights = panel_heights, widths = 30) 
```

```{r}
mysave(panel_b, name = "xgboost_beeswarm", width = 15, height = 10)
```

## Combining panels A, B and D

```{r}
gridExtra::arrangeGrob(panel_gain, panel_b, panel_harrell,
                       layout_matrix = rbind(c(1, 2), c(3, 3)),
                       widths = ggplot2::unit(c(90, 80), c("mm", "mm")),
                       heights = ggplot2::unit(c(120, 30), c("mm", "mm"))) %>%
  mysave(name = "xgboost", width = 7.5, height = 6.5)
```



## Partial dependecy plot

### Main features

```{r}
c_pdp_features <- c("BL_AGE", "SYSTM") %union%
  stringr::str_subset(c_importance_order, "GUT_|Phylum_|Class_|Order_|Family_|Genus_|Species_|NMR_")

c_pdp_features
c_importance_order <- "NMR_XL_VLDL_FC_pcnt"
```

### Data

```{r, eval = FALSE, echo = FALSE}
plot_data <- readRDS("rds/plot_data-20240326_160947.rds")
```


```{r}
plot_data <- nested_xgboost_cox_cv_pdp(results, c_pdp_features)
mysave(plot_data)
```

### Main plot

```{r}
factor_conversion <- rlang::expr(ifelse(stringr::str_detect(Feature, "GUT_|Phylum_|Class_|Order_|Family_|Genus_|Species_"), glue::glue("*{Desc}*"), as.character(Desc)))
```

```{r}
factor_order <- c("Age at baseline", "Systolic blood pressure") %union%
  { df_importance_sum %>% dplyr::pull(Desc) %>% unique %>% as.character %>% sort } %>%
  unique
```

```{r}
df_prevalence <- tses_gg2_MGS_FR02_nmr %>%
  purrr::map_df(~mia::getPrevalence(.x, detection = 0.1/100, as_relative = TRUE), .id = "rank") %>% 
  tidyr::gather(Feature, prevalence, -rank) %>%
  dplyr::mutate(Feature = stringr::str_replace(Feature, "Species:", "")) %>% 
  dplyr::mutate(Feature = glue::glue("{rank}_{Feature}")) %>% 
  tidyr::drop_na() %>% 
  dplyr::mutate(Feature = stringr::str_replace_all(Feature, c(" " = "_", ".*:" = ""))) %>%
  dplyr::left_join(df_importance_sum, by = dplyr::join_by(Feature)) %>%
  dplyr::mutate(Prevalence = glue::glue("Prevalence {sprintf('%.1f', prevalence*100)}%")) %>%
  my_df_factor_recode(Desc, factor_conversion) %>%
  dplyr::filter(Feature %in% c_importance_order)
```

```{r}
df_prevalence %>% DT::datatable()
```

### PDP plot

```{r}
df_pdp <- left_normalize_plot_data(plot_data, abundance, Feature, model) %>%
  dplyr::left_join(df_importance_sum, by = dplyr::join_by(Feature)) %>%
  dplyr::filter(Feature %in% c_pdp_features) %>%
  dplyr::filter(dplyr::n() == max(dplyr::row_number()), .by = c(Feature, model, yhat)) %>% 
  dplyr::mutate(Desc = forcats::fct_relevel(Desc, factor_order)) %>%
  my_df_factor_recode(Desc, factor_conversion)
```

```{r}
my_x_limits <- function(x) {
  diff <- max(x) - min(x)
  min_val <- { dplyr::case_when(min(x) < 10 & diff > 1 ~ floor(min(x)),
                                min(x) < 10 & diff < 1 ~ round(min(x)*20, digits = 0)/20,
                                TRUE ~ round(min(x)/5, digits = 0)*5)}

  max_val <- { dplyr::case_when(max(x) < 6 & diff > 1 ~ max(x),
                                max(x) < 11 & diff > 1 ~ 10,
                                max(x) < 16 & diff > 1 ~ 15,
                                max(x) < 10 & diff < 1 ~ round(max(x)*10, digits = 0)/10,
                                TRUE ~ round(max(x)/5, digits = 0)*5)}
    
  c(min_val, max_val)
}
```

```{r}
my_x_breaks <- function(x, limits = FALSE) {
  diff <- max(x) - min(x)
  step <- dplyr::case_when(diff < 4 ~ 0.2,
                           max(x) < 6 ~ 1,
                           diff > 4 & diff < 40 ~ 5,
                           diff > 40 & diff < 100 ~ 10,
                           TRUE ~ 20)

  seq(0, 300, step)
}
```

```{r}
df_data_for_rug <- df_gg2_MGS_FR02_nmr %>%
  dplyr::filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA), PREVAL_IHD == 0) %>%
  dplyr::select(one_of(c_pdp_features)) 
```

```{r}
df_rug <- df_pdp %>%
  dplyr::select(Feature, Desc) %>%
  unique %>%
  dplyr::left_join(df_prevalence %>% dplyr::select(Feature, prevalence), by = dplyr::join_by(Feature)) %>%
  dplyr::mutate(splits = dplyr::case_when(is.na(prevalence) ~ 1000,
                                          prevalence > 0.05 ~ ceiling(1000*prevalence),
                                          TRUE ~ 50)) %>% 
  dplyr::mutate(y_norm = purrr::map2(Feature, splits,
                                  ~stats::quantile(df_data_for_rug %>% dplyr::pull(.x), probs = seq(0,1,1/(.y-1))))) %>%
  tidyr::unnest(y_norm)
```

```{r}
df_rug %>% dplyr::group_by(Feature) %>% dplyr::summarize(n = dplyr::n())
```

```{r}
{ df_pdp %>%
    dplyr::filter(abs(yhat_norm) < 10) %>% 
    ggplot2::ggplot(ggplot2::aes(x = y_norm, y = yhat_norm)) +
    ggplot2::facet_wrap(~Desc, scales = "free", ncol = 3) + 
    ggplot2::geom_line(ggplot2::aes(group = model), alpha = 0.025) +
    ggplot2::geom_hline(yintercept = 0, color = "red") +
    ggplot2::geom_smooth(se = FALSE, colour="red", linetype="11", method = "gam", formula = y ~ s(x, bs = "cs")) +
    ggplot2::scale_color_grey(start = 0, end = 0.9) +
    ggplot2::geom_rug(data = df_rug,
                      ggplot2::aes(x = y_norm),
                      color = "red",
                      alpha = 0.1,
                      size = 1,
                      sides="b",
                      inherit.aes = FALSE) +
    ggplot2::guides(color = "none") +
    ggh4x::facetted_pos_scales(x = list(
                                 Desc == "Age at baseline" ~ ggplot2::scale_x_continuous(name = NULL,
                                                                                         breaks = seq(0, 100, 10)),
                                 Desc == "Systolic blood pressure" ~ ggplot2::scale_x_continuous(name = NULL,
                                                                                                 breaks = seq(0, 300, 20)),
                                 TRUE ~ ggplot2::scale_x_continuous(name = NULL,
                                                                    breaks = seq(-10, 10, 2)))) +
    ggplot2::scale_y_continuous(name = NULL,
                                breaks = scales::breaks_pretty(n = 5)) +
    ggplot2::coord_cartesian(ylim = c(-1, 1)) +
    ggplot2::geom_text(ggplot2::aes(label = Prevalence), x = -Inf, y = -Inf, hjust = -0.1, vjust = -1, data = df_prevalence) +
    ggplot2::theme(strip.text.x = ggtext::element_markdown(size = 13, hjust = 0),
                   panel.border = ggplot2::element_blank(),
                   plot.margin = ggplot2::margin(1, 5, 1, 1, "mm"), 
                   axis.line.x = ggplot2::element_line(size = 0.5, linetype = "solid", colour = "black"),
                   axis.line.y = ggplot2::element_line(size = 0.5, linetype = "solid", colour = "black")) } %>%
  mysave(name = "xgboost_pdp", width = 8, height = 8, dpi = 900)
```
